---
title: "Chad Huntebrinker Homework 11"
author: "Chad Huntebrinker"
date: "2025-04-21"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(VGAM)
library(readxl)

alligator_data <- read_excel("alligator_data.xlsx")

data <- data.frame(
  Treatment = factor(rep(c("Sequential", "Alternating"), each = 6)),
  Gender = factor(rep(rep(c("Male", "Female"), each = 3), 2)),
  Response = ordered(rep(c("NoChange", "Partial", "Complete"), 4),
                     levels = c("NoChange", "Partial", "Complete")),
  Count = c(45, 29, 26,
            12, 5, 2,
            44, 20, 20,
            7, 3, 1)
)
```


# Problem 6.2
For this problem, we need to use a model fit to estimate an odds ratio that describes the effect of length on primary food choice being either invertebrate or other. We will take the data and subset it so that it's just invertebrate or other in our data.
```{r echo=FALSE}
#We're subsetting the data to only keep I and O.
subset_data <- subset(alligator_data, y %in% c("I", "O"))

#Create the model
model <- vglm(y ~ x, data = subset_data, family = cumulative(parallel = TRUE, link = "logitlink"))
summary(model)

#Now we get the odds ratio
cat("Odds ratio:", exp(coef(model)["x"]), "\n")

#For each one-unit increase in length, the odds that the alligator chooses Invertebrate (vs Other) as
#its primary food decrease by about 89% (since 1 - 0.113 = 0.887).
```


For each one meter increase in length, the odds that the alligator chooses invertebrate as its primary food decrease by about 89% (since e^-2.179 = 0.113 and 1 - 0.113 = 0.887).

# Problem 6.8
We need to fit a cumulative logit model  with a proportional odds structure, interpret the estimated treatment effect, and check whether a model allowing interaction provides a significant better fit.
```{r echo=FALSE}
#Fit cumulative logit model
model_1 <- vglm(Response ~ Treatment + Gender,
                   family = cumulative(parallel = TRUE, link = "logitlink"),
                   weights = Count,
                   data = data)


summary(model_1)
#The logit model was fit to evaluate the effects of treatment type and gender on chemotherapy response.
#The estimated treatment effect comparing sequential to alternating therapy found the odds ratio
#as = 0.81 (exp(-0.2152) = 0.81) with a p-value of 0.41. The estimated treatment effect comparing
#the two genders had an odds ratio of 0.49 (exp(-0.7149) = 0.49) with a p-value of 0.07. This showed that
#sequential treatment had about 19% lower odds of having a better remission (complete remission
#being the best) and males had a 51% lower odds of having a better remission. However, since both of these
#estimations had a p-value < 0.05, neither one of them are significant to the response.
```
The logit model was fit to evaluate the effects of treatment type and gender on chemotherapy response. The estimated treatment effect comparing sequential to alternating therapy found the odds ratio of 0.81 (exp(-0.2152) = 0.81) with a p-value of 0.41. The estimated treatment effect comparing the two genders had an odds ratio of 0.49 (exp(-0.7149) = 0.49) with a p-value of 0.07. This showed that sequential treatment had about 19% lower odds of having a better remission (complete remission being the best) and males had a 51% lower odds of having a better remission. However, since both of these estimations had a p-value < 0.05, neither one of them are significant to the chemotherapy response.

Next, we'll see if a model with interaction fits better.
We'll do a Likelihood ratio test to see if the interaction model gives us a better fit. Our null hypothesis is that the model with interaction does not provide a better fit, while our alternative hypothesis is that the model with the interaction does.
```{r echo=FALSE}
#Next, we'll see if a model with interaction fits better.
model_2 <- vglm(Response ~ Treatment + Gender + Treatment:Gender,
                family = cumulative(parallel = TRUE, link = "logitlink"),
                weights = Count,
                data = data)

summary(model_2)

#We'll do a Likelihood ratio test to see if the interaction model gives us a better fit.
#Our null hypothesis is that the model with interaction does not provide a better fit, while our
#alternative hypothesis is that the model with the interaction does.
lrtest(model_1, model_2)
#With a p-value of 0.8, we can't reject our null hypothesis. So our model with an interaction term does not
#give us a better fit.
```
Our likelihood ratio test gives us a p-value of about 0.8. Since that is > 0.05, we can't reject our null hypothesis. So in conclusion, our model with an interaction term does not give us a better fit than the simpler model.

```{r}
#Code Appendix
#Chad Huntebrinker
#Problem 6.2
library(readxl)
library(VGAM)

#We need to use a model fit to estimate an odds ratio that describes the effect of length on primary food
#choice being either invertebrate or other.

alligator_data <- read_excel("alligator_data.xlsx")

#We're subsetting the data to only keep I and O.
subset_data <- subset(alligator_data, y %in% c("I", "O"))

#Create the model
model <- vglm(y ~ x, data = subset_data, family = cumulative(parallel = TRUE, link = "logitlink"))
summary(model)

#Now we get the odds ratio
cat("Odds ratio:", exp(coef(model)["x"]), "\n")

#For each one-unit increase in length, the odds that the alligator chooses Invertebrate (vs Other) as
#its primary food decrease by about 89% (since 1 - 0.113 = 0.887).

#Chad Huntebrinker
#Problem 6.8
#We need to fit a cumulative logit model  with a proportional odds structure, interpret the estimated
#treatment effect, and check whether a model allowing interaction provides a significant better fit.

library(VGAM)
library(readxl)

#Create the dataframe
data <- data.frame(
  Treatment = factor(rep(c("Sequential", "Alternating"), each = 6)),
  Gender = factor(rep(rep(c("Male", "Female"), each = 3), 2)),
  Response = ordered(rep(c("NoChange", "Partial", "Complete"), 4),
                     levels = c("NoChange", "Partial", "Complete")),
  Count = c(45, 29, 26,
            12, 5, 2,
            44, 20, 20,
            7, 3, 1)
)

#Fit cumulative logit model
model_1 <- vglm(Response ~ Treatment + Gender,
                   family = cumulative(parallel = TRUE, link = "logitlink"),
                   weights = Count,
                   data = data)


summary(model_1)
#The logit model was fit to evaluate the effects of treatment type and gender on chemotherapy response.
#The estimated treatment effect comparing Sequential to Alternating therapy found the odds ratio
#as = 0.81 (exp(-0.2152)) with a p-value of 0.41. The estimated treatment effect comparing
#the two genders had an odds ratio of 0.49 (exp(-0.7149)) with a p-value of 0.07. This showed that
#Sequential treatment had about 19% lower odds of having a better remission (complete remission
#being the best) and 51% lower odds of females having a better remission. However, since both of these
#estimations had a p-value < 0.05, neither one of them are significant to the response.

#Next, we'll see if a model with interaction fits better.
model_2 <- vglm(Response ~ Treatment + Gender + Treatment:Gender,
                family = cumulative(parallel = TRUE, link = "logitlink"),
                weights = Count,
                data = data)

summary(model_2)

#We'll do a Likelihood ratio test to see if the interaction model gives us a better fit.
#Our null hypothesis is that the model with interaction does not provide a better fit, while our
#alternative hypothesis is that the model with the interaction does.
lrtest(model_1, model_2)
#With a p-value of 0.8, we can't reject our null hypothesis. So our model with an interaction term does not
#give us a better fit.
```

